---
title: Mendelian errors in ALSPAC trios?
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: Gibran Hemani
---

Setup

```{r}
library(knitr)
knitr::opts_chunk$set(cache=TRUE)

library(tidyverse)
library(data.table)
library(glue)
library(jsonlite)

config <- read_json("config.json")
round1bfile <- config$round1bfile
round2bfile <- config$round2bfile
dir.create("data")
```

Read in data

```{r}
round1fam <- fread(paste0(round1bfile, ".fam"), header=FALSE) %>% as_tibble()
round2fam <- fread(paste0(round2bfile, ".fam"), header=FALSE) %>% as_tibble()
```

Any individuals present in both rounds?

```{r}
table(round1fam$V1 %in% round2fam$V1)
table(round2fam$V1 %in% round1fam$V1)
````

Merge round 1 and round 2 - just use chr22 to reduce computational time

```{r}
cmd <- glue("plink --bfile {round1bfile} --bmerge {round2bfile} --make-bed --out data/chr22_merge")
system(cmd)

cmd <- glue("plink --bfile {round1bfile} --exclude data/chr22_merge-merge.missnp --make-bed --out data/round1")
system(cmd)

cmd <- glue("plink --bfile {round2bfile} --exclude data/chr22_merge-merge.missnp --make-bed --out data/round2")
system(cmd)

cmd <- glue("plink --bfile data/round1 --bmerge data/round2 --make-bed --out data/chr22_merge")
system(cmd)

cmd <- glue("cp data/chr22_merge.fam data/chr22_merge.fam.orig")
system(cmd)
```

Read in the merged fam file

```{r}
mergefam <- fread("data/chr22_merge.fam.orig", header=FALSE) %>% as_tibble()
```

Create trios in the fam file

```{r}
kids <- grep("[A,B,C,D]", mergefam$V1, value=TRUE)
kidsaln <- gsub("[A-D]", "", kids)

mums <- grep("M", mergefam$V1, value=TRUE)
mumsaln <- gsub("M", "", mums)

dads <- grep("F", mergefam$V1, value=TRUE)
dadsaln <- gsub("F", "", dads)

kids_with_dads <- which(kidsaln %in% dadsaln)
length(kids_with_dads)

kids_with_mums <- which(kidsaln %in% mumsaln)
length(kids_with_mums)

mergefam$V3 <- as.character(mergefam$V3)
mergefam$V3[mergefam$V2 %in% kids][kids_with_dads] <- paste0(kidsaln[kids_with_dads], "F")

mergefam$V4 <- as.character(mergefam$V4)
mergefam$V4[mergefam$V2 %in% kids][kids_with_mums] <- paste0(kidsaln[kids_with_mums], "M")

mergefam$V1 <- gsub("[A-Z]", "", mergefam$V1)
write.table(mergefam, file="data/chr22_merge.fam", row=F, col=F, qu=F)
```

Get the allele frequencies of the merged file

```{r}
cmd <- glue("plink --bfile data/chr22_merge --freq --out data/chr22_merge")
system(cmd)

freq <- fread("data/chr22_merge.frq") %>% as_tibble()
table(freq$MAF > 0.05)
```

A lot of rare variants, restrict mendelian error analysis to only common variants

```{r}
cmd <- glue("plink --bfile data/chr22_merge --maf 0.05 --mendel --out data/chr22_merge")
system(cmd)
```

How many trios?

```{r}
ntrio <- sum(mergefam$V3 != "0" & mergefam$V4 != "0")
ntrio
```

How many variants

```{r}
nvar <- sum(freq$MAF > 0.05, na.rm=T)
nvar
```

Total number of possible errors

```{r}
nvar * ntrio
```

Actual number of errors


```{r}
cmd <- glue("cut -c 244 data/chr22_merge.mendel | sed 1d")
mendel <- system(cmd, intern=TRUE)
length(mendel)
```

ratio:

```{r}
length(mendel) / (nvar * ntrio)
```

Check distribution of errors

```{r}
imendel <- fread("data/chr22_merge.imendel") %>% as_tibble()
hist(imendel$N)
lmendel <- fread("data/chr22_merge.lmendel") %>% as_tibble()
hist(lmendel$N)
fmendel <- fread("data/chr22_merge.fmendel") %>% as_tibble()
hist(fmendel$N)
```

Relatedness - there's probably a faster way to do this but whatever

```{r}

trios <- mergefam$V3 !=0 & mergefam$V4 !=0
trioids <- bind_rows(
	tibble(V1=mergefam$V1[trios], V2=mergefam$V2[trios]),
	tibble(V1=mergefam$V1[trios], V2=mergefam$V3[trios]),
	tibble(V1=mergefam$V1[trios], V2=mergefam$V4[trios])
)

write.table(trioids, file="data/trioids.txt", row=F, col=F, qu=F)
cmd <- glue("plink --bfile data/chr22_merge --make-grm-gz --maf 0.05 --keep data/trioids.txt --out data/chr22_merge")
system(cmd)
rels <- fread("data/chr22_merge.grm.gz") %>% as_tibble()
relids <- fread("data/chr22_merge.grm.id") %>% as_tibble()

triofam <- mergefam[trios,]
triofam$pat <- triofam$mat <- NA

for(i in 1:nrow(triofam))
{
	i1 <- which(relids$V2 == triofam$V2[i])
	i2 <- which(relids$V2 == triofam$V3[i])
	i3 <- which(relids$V2 == triofam$V4[i])
	triofam$pat[i] <- rels$V4[rels$V2==i1 & rels$V1==i2]
	triofam$mat[i] <- rels$V4[rels$V2==i1 & rels$V1==i3]
}
```

All relationships:

```{r}
hist(rels$V4, breaks=100)
```

Just father-child:

```{r}
hist(triofam$pat)
```

Just mother-child:

```{r}
hist(triofam$mat)
```

